visitation_bias

### 2021-11-10 ms_1/response_to_reviewers.Rmd

---
title: "Response to reviewer comments"
output: pdf_document
# bibliography: 
  # - references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Editor's summary

## Summary

Both reviewers liked many aspects of the manuscript but also had strong suggestions for improving it, especially for clarifying methods and the Introduction.  The two reviewers different on how to approach the introduction and I find myself falling on the side of Reviewer 2. Both I and Reviewer 2 (perhaps stated a bit differently) agree that the framing of the story needs to be rethought some.  Reviewer 2 sees it as a more ecological story of plant community response driving pattern of bumble bee interactions. I agree that the story is really more about patterns along a gradient rather than a paper about climate change and potential responses. This will mean rethinking the introduction and how the paper is set up, but I think will make it a more robust match to the data and still use the strong analytical approaches you have.

The reviewers felt that the methods need more clarification in several places. I felt overall you did a solid job of explaining what are some relatively novel approaches and these show promise for other similar types of data from different systems.

Below I have a several general broader comments/suggestions and then also detail specific comments chronologically from my own review.

## General comments

The set up of the introduction leads with climate change, elevational shifts etc. however, the data set and results are not really about this, but rather about how differential responses of flowering vegetation communities versus bees across an environmental gradient can produce thresholds in interaction pattern along such gradients.  I agree the could be (likely will be) implications for communities’ responses to climate change.  However, because the data do not look at shifts and changing distributions the framing of the introduction is mismatched.  The data on ecological gradients, niches, and emergent interaction beta-diversity seem plenty solid to me and I think recasting with that as the main story would be more in line with the data, then moving on to consider possible implications for climate change. The climate change implications could be developed mostly in the discussion.

**COMMENT**
A related point, I very much like the points about elevational gradients and mountains as valuable “ecological theaters”. There must be other ecological situations where similar gradients and discontinuities exist: coastal intertidal, estuary salinity gradients? If you were able to find other examples, I think this would broaden the relevance with the general ecological audience of Ecology and strengthen the paper.

**RESPONSE**
Thank you for your constructive critique of our framing. We have rewritten our introduction accordingly, focusing on the intrinsic ecology of environmental gradients rather than climate-induced range shifts. We agree that this is both a more natural framing of our data and a richer ecological story.

**COMMENT**
In the methods I wanted more detail of the sampling over sites and time- you could include a table of effort among dates and years etc. for the sites, or you could provide more detail in the text.

**RESPONSE**
In our original submission, we placed the detailed description of sampling patterns in Appendix S1. Based on your feedback, we have moved the description of sampling patterns into the main text, summarized in Figure 2.

**COMMENT**
In looking at Figure 2, if I am interpreting the figure correctly the lines are modeled standardized abundance for each species over the elevation gradient. The patterns of peak bumble bee abundance among panels looks consistent with a generally uniform turnover of species across elevation.  Its almost like a sort of elevational niche partitioning. Isn’t this something Pyke wrote about for Rocky Mountain bumble bees.  This seems like a good point to bring up and discuss. I am not sure how closely the point ties to Figure 4, but it seems consistent with the smooth line for bumble bees in that figure.

**RESPONSE**
We have modified both the visualization (now Figure 5) and discussion of our bumble bee abundance model to more clearly depict the differences in elevational responses across species and explore its potential causes.

A paragraph in the revised discussion reads:

"The response of overall bumble bee abundance to elevation was approximately flat, but this aggregate pattern concealed divergent species-level responses. The strongly positive responses of B. mendax, B. monticola, B. mucidus, B. pyrenaeus are consistent with the known affinity of these species for high-elevation habitats (Rasmont and Iserbyt 2010-2014). At the other extreme, B. pascuorum and the B. psithyrus species group exhibited strongly negative responses to elevation. In the case of B. pascuorum, this pattern is likely driven by its exceptionally long nesting cycle, which limits its fitness in the short growing seasons of higher elevations (Neumayer 2009). Growing season length may also explain the lower abundance of B. psithyrus at high elevation, since cuckoo bumble bees must emerge later than the queens of their host species and then successfully usurp host nests before they can begin to raise brood (Goulson 2010). Nevertheless, while the divergent responses of bumble bee species to elevation are informative, it is important to emphasize the overall robustness of bumble bees to elevational variation. Nine out of the 16 species in our study system did not respond significantly to elevation, and 8 out of 16 were observed across the full span of the gradient."

And we extend our discussion of these patterns to explore the question of why bubmle bees forage in resource-poor high-elevation habitats:

"The poverty of floral resources at the highest elevations raises the question of why bumble bees choose to forage at such elevations at all. For some species, the selection of high elevation habitats may be influenced by thermal tolerances (Oyen et al. 2016), but that fails to explain the presence of elevation-generalist species (including the majority of the species in our study system) in resource-poor high-elevation habitats. It is possible that nest site limitation at lower elevations drives some bumble bees to establish nests at high elevation, and their ability to forage downslope is subsequently constrained by their flight range, which may be reduced due to the energetic costs of flight in cold environments (Kovac et al. 2015). It is also possible that high elevation landscapes, while sparse in absolute floral resources, offer an escape from more competitive conditions at lower elevations where forager density is higher (Hoiss et al. 2015, Classen et al. 2020). Similarly, higher elevations could also offer an escape from density-dependent parasite and pathogen pressure (Parsche and Lattorff 2018). Finally, it is worth speculating that an affinity for open, high-elevation habitats might be an evolutionary legacy in mountain bumble bees, since the existence of flower-rich meadows in the forest matrix below the tree line is largely an artifact of human land use arising mainly in the last 500 years of bumble bees’ evolutionary history (Hejcman et al. 2013)."

## Specific comments

**COMMENT**:
Line 22  add “many of which”  somewhere around associated with mountains to describe bumble bees

**RESPONSE**:
Our revised abstract no longer contains this wording.

**COMMENT**
Line 33-34 this wording is awkward “to demonstrate”- doesn’t seem like quite the right word choice.

**RESPONSE**
Our revised abstract no longer contains this wording.

**COMMENT**
Line 46 abstract translations – wording here again seems odd.

**RESPONSE**
Our revised introduction no longer contains this wording.

**COMMENT**
Line 47-49 break into two sentences  this would improve clarity.  New sentence after the Blois et al citation. “The elevational gradients also involve variation in abiotic conditions other than temperature.”

**RESPONSE**
Our revised introduction no longer contains this wording.

**COMMENT**:
Line 60 delete species after cold-adapted.  As worded it implies that you are viewing bumble bees as a species.

**RESPONSE**
Section now reads: "Today, they remain characteristically cold-adapted organisms (Heinrich and Esch 1994)"

**COMMENT**
Line 71 replace the with their after hinge upon and delete of bumble bees after elevational distributions.

**RESPONSE**
Our revised introduction no longer contains this wording.

**COMMENT**
Line 80 I suggest that you spell out more clearly what you mean by geometric considerations, perhaps “area contractions”?

**RESPONSE**
Our revised introduction no longer contains this wording.

### Methods

**COMMENT**
I was struck by potentially unequal effort among sites.  I recognize that sampling within a site visit was consistent but different numbers of visits to sites within and among years would presumably create potential bias in detection among species and affect measures of abundance and richness.  These in turn could affect b-diversity. It would be helpful to see some coverage of this point including a reference to why it is not an issue (if indeed it isn’t).

**RESPONSE**
As described above, we have moved the description of sampling patterns into the main text, summarized in Figure 2. We account for the potential effect of sampling intensity (i.e. the total number of sampling transects per site) on observed beta-diversity by including sampling intensity as a covariate in our GDM analysis. The relevant line of the methods section reads:

"To control for potential confounding effects of sampling intensity, sampling days per site was included as a covariate."

**COMMENT**
Line 138 suggest changing sparsity of observations to sparse observations

**RESPONSE**
Our revised method section no longer contains this wording.

**COMMENT**
Lines 141-144.  Rarity alone doesn’t seem like a justification for omission.  These taxa would have little weight in such an assessment and so don’t need removing for that reason alone. If the issue was due to model fitting or something then I would suggest justifying it what way.  Otherwise, it seems odd to prioritize their inclusion in beta-diversity calculation but not elsewhere.

**RESPONSE**
In our revised manuscript, all models have been refit to include the previously omitted species. The main reason for omitting them in our initial analysis was that it was difficult to model species-specific floral resource availability for rare species. During our revision, we reconsidered the value of that analysis, since species did not exhibit strong differences in floral resource patterns anyway, and it added considerable complexity to our models. In our revised analysis, we model only aggregate floral resource availability, and we include all bumble bee species in our models of bumble bee richness, abundance, and beta-diversity.

Note that we have also opted to model peak rather than mean floral abundance. This decision did not change the resulting patterns very much, but we think as a matter of principle that modeling the mean of a log-normal distribution is problematic, and peak abundance seems a more natural summary for an essentially unimodal phenological process.

**COMMENT**
Line 143  write out genus name at the start of sentence.

**RESPONSE**
Done.

**COMMENT**
Line 182 This seems like a cool thing to look at and a powerful tool (although I am not really familiar with the approach).  It made me wonder why something similar wasn’t done to look at overall patterns of abundance and diversity for bees and flowers.

**RESPONSE**
In our revised manuscript, we include models of flower and bumble bee species richness. We opt for GAM analysis of these patterns, though, rather than GDM analysis. We chose GDM analysis of beta-diversity because beta-diversity is inherently a measure of change, and therefore best characterized in terms of rates.

**COMMENT**
One note about the beta-div partitioning analysis.  I believe that bi-partite package now has a way to do this quantitatively (new paper out by Jochen Freund), something you could consider, but I was also fine with it as is. You probably have seen this too.

**RESPONSE**
The paper you reference describes the function (and settings thereof) that we used in our analysis. The paper may have just come out, but it seems the method has been implemented in `bipartite` since the 2.14 release in early 2020.

**COMMENT**
Line 241 grazing in American English? I expect it would get changed in copy editing.

**RESPONSE**
We have replaced "grasing" with "grazing" throughout the manuscript.

**COMMENT**
Line 282 This sentence is awkward because its hard to know what you found differences in =peak abundance.  Also the next sentence is a pretty interesting statement, but I don’t think is quantitatively explored in the paper.  It would strengthen the argument about the special relevance of the tree line to do so. We expect species that are decline and those that are increase in abundance to have overlap in the middle, so not surprising, but your claim is interesting and more specific. Is there a way to assess if the tree line is indeed a particular zone.  It seems tricky to isolate because species actually appear to peak at different elevations so the slopes of their relative abundance will be different at each point in elevation.  Perhaps if the tree line was a point of highest absolute value in slopes among species or something?

**RESPONSE**
During revisions, we reconsidered our initial emphasis on the tree line ecotone, and we concluded that it was over-extended in the case of bumble bees (though not in the case of plants). In our revised manuscript, we do not describe the tree line ecotone as a special zone of turnover for bumble bees, and such an interpretation is not supported in our revised GDM analysis. We do, however, note that the tree line ecotone seems to be a significant peak of *B. pratorum* abundance. This section now reads:

"At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation (Figure 5; Table 3). Bombus pyrenaeus, B. mendax, B. monticola, and B. mucidus exhibited approximately linear increases with elevation. In contrast, Bombus psithyrus and B. pascuorum exhibited approximately linear decreases with elevation. Bombus pratorum was unique in exhibiting a strongly unimodal abundance pattern with a well-defined peak between 1500-1600 m, approximately at the tree line ecotone, similar to the abundance pattern of B. pratorum described by Iserbyt et al. (2008) in the Eyne Valley of the Pyrenees. The remaining species did not respond significantly to elevation (p > 0.05)."

**COMMENT**
Line 290.  You say its notable that for some species one peak of flower resources falls outside their elevational zone.  Why would we expect it to be otherwise?  It would only be otherwise if for some reason bumble bee species were driving the distribution of their floral hosts.  Can you develop the point more clearly?

**RESPONSE**
In our revised manuscript, we decided to drop the species-specific model of floral resource availability, so this statement no longer appears in the paper.

**COMMENT**
Line 320 This is an interesting result but I wonder if it could be anything otherwise, so that the story is really about an elevational shift in plant distribution and turnover with bumble bees interactions following.  The bees themselves do NOT actually that obviously follow the same sort of discontinuity.  The paper sort of implies they do and that this tree line is a key place, but as I mentioned in the comment (line 282) that is not really demonstrated from abundance distributions (only 1 of 12 species shows such a pattern) and would be good to somehow quantify in the case of the bumble bees themselves.

**RESPONSE**
As described in our response to a comment above, we have stepped back from the emphasis on tree-line effect with respect to bumble bees. As you point out, it is actually more interesting to note the *differences* in response between plants and bumble bees, and how change in the latter are best understood as responses to changes in the former. This change of perspective is reflected in this paragraph of our discussion section:

"The roughly sigmoidal pattern of bumble bee beta-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects both the reciprocal abundance patterns of high- and low-elevation species and the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral beta-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work (e.g. Pellissier et al. 2010, Descombes et al. 2017), and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium of species composition, but rather a threshold marking the beginning of a process of beta-diversity acceleration that continued for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m (Körner 2003), and the accelerated beta-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance (Figures 3B, 4B)."

...and we pick up the again when we discuss the implications of our findings for the response of mountain bumble bees and wildflowers to climate and land use change:

"Mountains are unique ecological theaters in which the extrinsic forces of climate and land use change interact with intrinsic environmental gradation to shape the ecology of species and their interactions (Peters et al. 2019). Given the elevational patterns of richness, abundance, and beta-diversity among bumble bees and wildflowers in our study, what can be said about the response of these communities, and their functional relationships, to climate and land use change? If the process of mid-elevation pasture abandonment continues, the flower-rich meadows that now perforate the forest matrix will eventually close into flower-poor coniferous forest, and bumble bees will be pushed either into the more agriculturally intensified land at lower elevation or the relatively resource-poor open habitat above the tree line. This land- use-driven redistribution of bumble bees will interact with the concomitant process of climate warming, which has already generated measurable upslope shifts in some bumble bees species (Ploquin et al. 2013, Pyke et al. 2016, Marshall et al. 2020), sometimes resulting in increased species richness at higher elevations (Fourcade et al. 2019). Under this scenario, high-elevation bumble bees can be expected to face increased densities of low-elevation or elevation-generalist species, and the introduction of competing bumble bee species in a given habitat can have strong effects both on the fitness of their congeners and on patterns of floral visitation (Nagamitsu et al. 2010, Ishii 2013, Brosi and Briggs 2013). Nevertheless, the tolerance of bumble bees for a relatively broad climate envelope together with their long flight range may enable them to cope with climate and land use change better than their floral mutualists. The flowering plant community in our study system was strongly structured by elevation, and there is evidence that the upslope movement of plants in response to warming temperatures may outpace that of bumble bees (Marshall et al. 2020, though see Pyke et al. 2016). Moreover, it is the plant community that experiences the direct effects of land use change. Thus, it may be that the strongest effects of climate and land use change on mountain bumble bees will be indirect effects mediated by changes in the floral community."

**COMMENT**
Lines 327. This is an interesting point, but doesn’t seem unique to bumble bees.  Presumably other generalist bee species face similar issues in montane areas as would other mountain species with varied resources.  This would be worth addressing.  Also it seems a stretch to attribute this montane effect to bumble bees in general unless the speculation is that mountain environments have played a key role in the evolution of bumble bees as a group.  Are there data to support that point?

**RESPONSE**
Mountains have played a central role in the origin and diversification of *Bombus*, so we think it is wortwhile to point out the behavioral challenge that this entails. To strengthen this point, we have added to our introduction an expanded discussion of the role of mountains in bumble bee evolution:

"Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods (Hines 2008, Stewart et al. 2010, Martinet et al. 2018). Today, they remain characteristically cold-adapted organisms (Heinrich and Esch 1994), with peak abundance and diversity in mountain ranges and northern latitudes (Williams 1998). In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora (Goulson 2010). As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures (Kerr et al. 2015, Soroye et al. 2020, though see Guzman et al. 2021). At the same time, socioeconomic drivers are causing the reforestation of pastures on mountain slopes and the intensification of agriculture in lowlands, both of which entail a loss of floral resources for bumble bees (Tasser and Tappeiner 2002, Tappeiner et al. 2006, Tasser et al. 2007, Walcher et al. 2017)."

**COMMENT**
The development of the implications of the research for climate change works better here in the discussion where it builds from the core findings, rather than as it is currently in the Introduction where it seems to motivate the research, but really isn’t targeted to address elevational shifts of bees, plants, etc.

**RESPONSE**
In the revised manuscript, our discussion of climate (and land use) change has been shifted mostly to the penultimate paragraph of the discussion (beginning with "Mountains are unique ecological theaters . . ."), which we quote in response to a previous comment.

\newpage

# Reviewer 1

## Summary

### 2021-11-10 code/ms2_nullmodels.Rmd
push to cloud just in case

## 2.1 Observed vs. null predictions
```{r}
# Species-wise
ggplot(null_tab1, aes(Observed, ktype)) +
  geom_segment(aes(y = ktype, yend = ktype, 
                   x = Lower.99.CL, xend = Upper.99.CL), 
               color = "gray40", inherit.aes = FALSE) +
  geom_point(aes(color = Test)) +
  facet_wrap(~fct_rev(bb.sp)) +
  scale_y_discrete(limits = rev) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  labs(x = "Mean visitation rate", y = "Floral morphotype", color = "Test (p < 0.01)") +
  theme_light()

@@

### 2.4.5 Cluster preferences
```{r}
clust1 <- null1_prefs %>%
  vegdist(method = "euclidean") %>%
  hclust(method = "complete")

>>>>

## 2.1 Observed vs. null predictions
```{r}
# Species-wise
ggplot(null_tab1, aes(Observed, ktype)) +
  geom_segment(aes(y = ktype, yend = ktype, 
                   x = Lower.99.CL, xend = Upper.99.CL), 
               color = "gray40", inherit.aes = FALSE) +
  geom_point(aes(color = Test)) +
  #facet_wrap(~fct_rev(bb.sp)) +
  facet_wrap(~bb.sp) +
  scale_y_discrete(limits = rev) +
  scale_color_manual(values = c("gray40", "#F8766D", "#00B6EB")) +
  labs(x = "Mean visitation rate", y = "Floral morphotype", color = "Test (p < 0.01)") +
  theme_light()

@@

### 2.4.5 Cluster preferences
```{r}
clust1 <- null1_prefs %>%
  vegdist(method = "euclidean") %>%
  hclust(method = "ward.D2")
